A Linear Solution of the Linearity Challenge

### 2025-12-26 paper/Context.lagda.tex
global revision

We are going to adopt a nameless representation of channels. Hence, typing
contexts are represented as lists of types, where the (polymorphic)
\AgdaDatatype{List} type and its constructors \AgdaInductiveConstructor{[]} and
\AgdaInductiveConstructor{\_∷\_} are defined in the module
\AgdaModule{Data.List} of Agda's standard library. We will keep using
$\ContextC$, $\ContextD$ and $\ContextE$ to range over typing contexts, even
though in the Agda formalisation they are lists and not finite maps as in
\Cref{sec:typing-rules}.

@@

Note that, when specifying a splitting $\Splitting\ContextC{[A]}\ContextE$ where
the left partition is a singleton $[A]$, it may be convenient to use
\AgdaFunction{≫} as a shortcut for a sequence of applications of
\AgdaInductiveConstructor{>} once the $A$ type has been reached in $\ContextC$.
For instance, \AgdaFunction{splitting-example₁} can be written equivalently and
in a more compact way as shown below. More usages of \AgdaFunction{≫} will be
provided in \Cref{sec:examples}.

@@

Hereafter we will make extensive use of predicates over contexts. For this
reason, it is worth recalling a number of definitions from the
\AgdaModule{Relation.Unary} module of Agda's standard library in
\Cref{tab:predicates}.

Context splitting is pervasive in the subsequent development. For this reason,
we define a few abstractions that allow us to hide context splittings and
enhance code readability, at least in some cases.
%
Inspired by separation logic~\cite{OHearnPym99} and
following~\citet{RouvoetPoulsenKrebbersVisser20}, we define the \emph{separating
conjunction} $P \sand Q$ of two predicates $P$ and $Q$ over contexts:
\begin{code}
data _∗_ (P Q : Pred Context _) (Γ : Context) : Set where
  _⟨_⟩_ : ∀{Δ Θ} → P Δ → Γ ≃ Δ + Θ → Q Θ → (P ∗ Q) Γ
\end{code}

@@

It is easy to prove that $\Splitting\ContextC\ContextD\ContextE$ is unrestricted
if so are $\ContextD$ and $\ContextE$:

>>>>

We are going to adopt a nameless representation of channels. Hence, typing
contexts are represented as lists of types, where the (polymorphic) type
\AgdaDatatype{List} and its constructors \AgdaInductiveConstructor{[]} and
\AgdaInductiveConstructor{\_∷\_} are defined in the module
\AgdaModule{Data.List} of Agda's standard library. We will keep using
$\ContextC$, $\ContextD$ and $\ContextE$ to range over typing contexts, even
though in the Agda formalisation they are lists and not finite maps as in
\Cref{sec:typing-rules}.

@@

When proving a splitting $\Splitting\ContextC{[A]}\ContextE$ where the left
partition is a singleton $[A]$, it may be convenient to use \AgdaFunction{≫} as
a shortcut for a sequence of applications of \AgdaInductiveConstructor{>} once
the $A$ type has been reached in $\ContextC$. For instance,
\AgdaFunction{splitting-example₁} can be written equivalently and in a more
compact way as shown below. More usages of \AgdaFunction{≫} will be provided in
\Cref{sec:examples}.

@@

From now on we will make extensive use of predicates over contexts. For this
reason, it is worth recalling in \Cref{tab:predicates} a number of definitions
from the module \AgdaModule{Relation.Unary} of Agda's standard library.
%
We begin using these definitions for building a few abstractions inspired to
separation logic~\cite{OHearnPym99} that allow us to hide context splittings, at
least in some cases.
%
Following~\citet{RouvoetPoulsenKrebbersVisser20}, we define the \emph{separating
conjunction} $P \sand Q$ of two predicates $P$ and $Q$ over contexts:
\begin{code}
data _∗_ (P Q : Pred Context _) (Γ : Context) : Set where
  _⟨_⟩_ : ∀{Δ Θ} → P Δ → Γ ≃ Δ + Θ → Q Θ → (P ∗ Q) Γ
\end{code}

@@

It is easy to prove that $\ContextC$ is unrestricted if so are $\ContextD$ and
$\ContextE$ when $\Splitting\ContextC\ContextD\ContextE$:

### 2025-12-26 paper/Permutations.lagda.tex
global revision

According to our nameless representation of channels, it is the \emph{position}
of a type in a typing context $\Context$ that determines the location of its
binder in the structure of a process. When the binding structure of a process
changes, \eg because a structural pre-congruence rule is applied, or when a
channel substitution occurs, \cf the right-hand side of the \LinkRule reduction
in \Cref{tab:semantics}, the typing context must be suitably rearranged. Such
rearrangement is in fact a \emph{permutation} of the elements of $\Context$.

@@

Just like permutations in Agda's standard library, also \AgdaDatatype{\_↭\_}
enjoys a number of expected properties, such as symmetry. In the following we
will make use of a specific property that relates context splitting, list
concatenation \AgdaFunction{\_++\_} and permutations: if
$\Splitting\ContextC\ContextD\ContextE$, then $\ContextC$ is a permutation of
the concatenation of $\ContextD$ and $\ContextE$.

>>>>

According to our nameless representation of channels, it is the \emph{position}
of a type in a typing context $\Context$ that determines the location of its
binder in the structure of a process. When the binding structure of a process
changes, \eg because a structural pre-congruence rule is applied, or when a
channel substitution occurs, \cf the right-hand side of the \LinkRule reduction
in \Cref{tab:semantics}, $\Context$ must be suitably rearranged to agree with
the updated binding structure. Such rearrangement is in fact a
\emph{permutation} of the elements of $\Context$.

@@

It is easy to see that \AgdaDatatype{\_↭\_} is an equivalence relation. In the
following we also use another specific property of permutations related to
context splitting and list concatenation \AgdaFunction{\_++\_}: if
$\Splitting\ContextC\ContextD\ContextE$, then $\ContextC$ is a permutation of
the concatenation of $\ContextD$ and $\ContextE$.

### 2025-12-26 paper/Process.lagda.tex
global revision

Channels are not given any name. Instead, they are represented as terms
witnessing that their type is present in the typing context. This is known as
\emph{co-de Bruijn syntax}~\cite{McBride18}, according to which the typing
context associated with a term (a process, a channel) only contains the types of
the channels that actually occur within the term. For this reason, typing
contexts are split eagerly, according to the structure of processes, to make
sure that channels are appropriately (and above all linearly) distributed among
sub-processes, so that each channel is used exactly once.
%
Concretely, a channel of type $A$ is a predicate that holds for the singleton
context $\AgdaFunction[~A~\AgdaFunction]$:

@@

The constructor \AgdaInductiveConstructor{link} models a process of the form
$\Link\x\y$. This process is well typed in a context of the form $x : A, y :
\dual{A}$, namely a context satisfying the predicate
$\AgdaDatatype{Ch}~A~\AgdaDatatype{∗}~\AgdaDatatype{Ch}~(\AgdaFunction{dual}~A)$
which we see on the left-hand side of \AgdaFunction{⇒}.
%
The constructor \AgdaInductiveConstructor{cut} models a cut $\Cut[A]\x{P}{Q}$.
For this process to be well typed in a context $\Context$, it must be the case
that $\Splitting\Context\ContextD\ContextE$ so that $P$ and $Q$ are well typed
in the contexts $x : A, \ContextD$ and $x : \dual{A}, \ContextE$ respectively.
Since $x$ is the most recently introduced channel, the types $A$ and $\dual{A}$
is added \emph{at the front} of $\ContextD$ and $\ContextE$, which we do by
means of the functions $(A~\AgdaInductiveConstructor{∷\_})$ and
$(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷\_})$ (these are partial
applications of the constructor $\AgdaInductiveConstructor{\_∷\_}$ for lists to
which we have supplied the left operand).

@@

For example, the constructor \AgdaInductiveConstructor{fail}, which corresponds
to a process of the form $\Fail\x$, requires the context $\Context$ to satisfy
the predicate
$\AgdaDatatype{Ch}~\AgdaInductiveConstructor{⊤}~\AgdaDatatype{∗}~\AgdaFunction{U}$,
meaning that $\Context$ must contain an entry \AgdaInductiveConstructor{⊤}
(\AgdaFunction{U} is the universal predicate that holds for every context, \cf
\Cref{tab:predicates}). That form of $\Context$ matches the typing context in
the conclusion of the rule \FailRule.

The constructor \AgdaInductiveConstructor{wait} corresponds to a process of the
form $\Wait\x.P$ and requires $\Context$ to (separately) satisfy
$\AgdaDatatype{Ch}~\AgdaInductiveConstructor{⊥}$, that is the channel $x$ on
which the process is operating, as well as \AgdaDatatype{Proc}, that is the
continuation process $P$, which must be well typed in the remaining typing
context.

The constructor \AgdaInductiveConstructor{close}, which corresponds to a process
of the form $\Close\x$, requires the typing context to be the singleton list
$[\One]$.

Let us move on to the processes that produce a continuation channel. For
example, the constructor \AgdaInductiveConstructor{case} corresponds to a
process of the form $\Case\x\y{P}{Q}$, where both $P$ and $Q$ use the
continuation channel $y$. In this case $\Context$ must satisfy the predicate
\[
  \AgdaDatatype{Ch}~(A~\AgdaInductiveConstructor{\&}~B)~\AgdaDatatype{∗}~((A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}~\AgdaFunction{∩}~(B~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc})
\]
which looks intimidating at first but makes perfect sense once we recall the
typing rule \CaseRule and the definitions of \AgdaFunction{∩} and
\AgdaFunction{⊢} in \Cref{tab:predicates}. Remember that we are trying to
establish whether $\Case\x\y{P}{Q}$ is well typed in $\Context$.
%
The predicate $\AgdaDatatype{Ch}~(A~\AgdaInductiveConstructor{\&}~B)$ expresses
the requirement that the type of $x$ must be of the form $A \With B$ and should
be found in $\Context$. In other words, it must be the case that
$\Splitting\Context{[A \With B]}\ContextD$ for some $\ContextD$. Now $P$ and $Q$
must be well typed in the context $\ContextD$ augmented with the association $y
: A$ and $y : B$, respectively, whence the use of \AgdaFunction{∩} to verify a
(non-separating) conjunction of the predicates
$(A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}$ and
$(B~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}$. The
two new contexts are obtained by \emph{adding} either $A$ or $B$ to $\ContextD$,
which we perform using \AgdaFunction{⊢}.
%
Crucially, the types $A$ and $B$ are \emph{prepended} to $\ContextD$, which is
consistent with the fact that the continuation $y$ has been freshly introduced
and the channel $x$ has been consumed. Notice how easy it is to \emph{prepend}
either $A$ or $B$ to $\ContextD$ instead of \emph{changing} the type of $x$ in
$\Context$ from $A \With B$ to either $A$ or $B$ while preserving the position
of the type, as we would have to do in a ``true'' session type system without
explicit continuation channels.

The interpretation of the remaining constructors is analogous, so we only
comment \AgdaInductiveConstructor{all}, which represents a polymorphic process
$\Forall\x\X\y.P$. This constructor models the continuation $P$ using
higher-order abstract syntax: witnessing that a context $\Context$ satisfies the
predicate
\[
  \AgdaFunction{⋂[}~X~\AgdaFunction{∶}~\AgdaDatatype{Type}~\AgdaFunction]~((\AgdaFunction{subst~[}~X~\AgdaFunction{/]}~A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}
\]
means providing a function that, for every type $X$, produces a witness for the
predicate
\[
  ((\AgdaFunction{subst~[}~X~\AgdaFunction{/]}~A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}
\]
applied to $\Context$. Note that the side condition $X \not\in \fv\Context$ of
\ForallRule is enforced \emph{by definition}:
%
$A$ has type $\AgdaDatatype{PreType}~\AgdaInductiveConstructor{1}$, that is a
pre-type with \emph{at most one} free type variable $X$, whereas $\Context$ is a
typing context, that is a list of $\AgdaDatatype{Type} =
\AgdaDatatype{PreType}~\AgdaInductiveConstructor{0}$ without free type
variables. Therefore, $X$ cannot occur in $\Context$.

>>>>

Channels are not given any name. Instead, they are represented as terms
witnessing that their type is present in the typing context. This is known as
\emph{co-de Bruijn syntax}~\cite{McBride18}, whereby the typing context
associated with a term (a process, a channel) only contains the types of the
channels that actually occur within the term. For this reason, typing contexts
are split eagerly, according to the structure of processes, to make sure that
channels are appropriately (and above all linearly) distributed among
sub-processes, so that each channel is used exactly once.
%
Concretely, a channel of type $A$ is a predicate that holds for the singleton
context $\AgdaFunction[~A~\AgdaFunction]$:

@@

The constructor \AgdaInductiveConstructor{link} builds a link $\Link\x\y$. This
process is well typed in a context of the form $x : A, y : \dual{A}$, namely a
context satisfying the predicate
$\AgdaDatatype{Ch}~A~\AgdaDatatype{∗}~\AgdaDatatype{Ch}~(\AgdaFunction{dual}~A)$
which we see on the left-hand side of \AgdaFunction{⇒}.

The constructor \AgdaInductiveConstructor{cut} builds a cut $\Cut[A]\x{P}{Q}$.
This process is well typed in a context $\Context$ if
$\Splitting\Context\ContextD\ContextE$ so that $P$ and $Q$ are well typed in the
contexts $x : A, \ContextD$ and $x : \dual{A}, \ContextE$, which are obtained
from $\ContextD$ and $\ContextE$ by adding the bindings $x : A$ and $x :
\dual{A}$, respectively. Since $x$ is the most recently introduced channel, the
types $A$ and $\dual{A}$ are added \emph{in front} of $\ContextD$ and
$\ContextE$, which we do by means of the functions
$(A~\AgdaInductiveConstructor{∷\_})$ and
$(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷\_})$. These are partial
applications of the constructor $\AgdaInductiveConstructor{\_∷\_}$ for lists to
which we have supplied the left operand.

@@

For example, the constructor \AgdaInductiveConstructor{fail}, which builds a
process $\Fail\x$, requires the context $\Context$ to satisfy the predicate
$\AgdaDatatype{Ch}~\AgdaInductiveConstructor{⊤}~\AgdaDatatype{∗}~\AgdaFunction{U}$,
meaning that $\Context$ must contain an entry \AgdaInductiveConstructor{⊤}
(\AgdaFunction{U} is the universal predicate that holds for every context, \cf
\Cref{tab:predicates}). That form of $\Context$ matches the typing context in
the conclusion of the rule \FailRule.

The constructor \AgdaInductiveConstructor{wait}, which builds a process
$\Wait\x.P$, requires $\Context$ to (separately) satisfy
$\AgdaDatatype{Ch}~\AgdaInductiveConstructor{⊥}$, that is the channel $x$ on
which the process is operating, as well as \AgdaDatatype{Proc}, that is the
continuation process $P$, which must be well typed in the remaining typing
context.

The constructor \AgdaInductiveConstructor{close}, which builds a process
$\Close\x$, requires the typing context to be the singleton list $[\One]$.

Let us move on to the forms that produce continuation channels. As an example,
the constructor \AgdaInductiveConstructor{case} builds a process
$\Case\x\y{P}{Q}$, where both $P$ and $Q$ use the continuation channel $y$. In
this case $\Context$ must satisfy the predicate
\[
  \AgdaDatatype{Ch}~(A~\AgdaInductiveConstructor{\&}~B)~\AgdaDatatype{∗}~((A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}~\AgdaFunction{∩}~(B~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc})
\]
which looks intimidating at first but makes perfect sense once we recall the
typing rule \CaseRule and the definitions of \AgdaFunction{∩} and
\AgdaFunction{⊢} in \Cref{tab:predicates}. Remember that we are trying to
establish whether $\Case\x\y{P}{Q}$ is well typed in $\Context$.
%
The predicate $\AgdaDatatype{Ch}~(A~\AgdaInductiveConstructor{\&}~B)$ expresses
the requirement that the type of $x$ must be of the form $A \With B$ and should
be found in $\Context$. In other words, $\Splitting\Context{[A \With
B]}\ContextD$ for some $\ContextD$. Now $P$ and $Q$ must be well typed in the
context $\ContextD$ augmented with the association $y : A$ and $y : B$,
respectively, whence the use of \AgdaFunction{∩} to verify a (non-separating)
conjunction of the predicates
$(A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}$ and
$(B~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}$. The
two new contexts are obtained by \emph{adding} either $A$ or $B$ to $\ContextD$,
which we perform using \AgdaFunction{⊢}.
%
Crucially, the types $A$ and $B$ are \emph{prepended} to $\ContextD$, which is
consistent with the fact that the continuation $y$ has been freshly introduced
and the channel $x$ has been consumed. Notice how easy it is to \emph{prepend}
either $A$ or $B$ to $\ContextD$ instead of \emph{changing} the type of $x$ in
$\Context$ from $A \With B$ to either $A$ or $B$ while preserving the position
of the type, as we would have to do in a ``true'' session type system without
explicit continuation channels.

The interpretation of the remaining constructors is analogous, so we only
comment \AgdaInductiveConstructor{all}, which builds a process
$\Forall\x\X\y.P$. This constructor models the continuation $P$ using
higher-order abstract syntax: proving that a context $\Context$ satisfies the
predicate
\[
  \AgdaFunction{⋂[}~X~\AgdaFunction{∶}~\AgdaDatatype{Type}~\AgdaFunction]~((\AgdaFunction{subst~[}~X~\AgdaFunction{/]}~A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}
\]
means providing a function that, for every type $X$, produces a witness for the
predicate
\[
  ((\AgdaFunction{subst~[}~X~\AgdaFunction{/]}~A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}
\]
applied to $\Context$. Note that the side condition $X \not\in \fv\Context$ of
\ForallRule is trivially enforced \emph{by definition}:
%
$A$ has type $\AgdaDatatype{PreType}~\AgdaInductiveConstructor{1}$, that is a
pre-type with \emph{at most one} free type variable $X$, whereas $\Context$ is a
typing context, that is a list of $\AgdaDatatype{Type} =
\AgdaDatatype{PreType}~\AgdaInductiveConstructor{0}$ without free type
variables. Therefore, $X$ cannot occur in $\Context$.

### 2025-12-26 paper/Type.lagda.tex
global revision

It is straightforward to prove that duality is a an involution.

@@

This property is important in the rest of the formalisation so we define an
implicit rewriting rule that Agda can autonomously apply whenever possible. This
is achieved by means of the following pragma directive.\footnote{The directive
is effective provided that the option \texttt{{-}{-}rewriting} is enabled,
either globally when invoking Agda or within a pragma in the module's source
code.}

@@

Next we define the function \AgdaFunction{subst} that simultaneously substitutes
the type variables of a pre-type with other pre-types. In practice we will
always substitute one variable at a time, but it is technically easier to
substitute \emph{all} variables of a pre-type, possibly with themselves. The
definition of \AgdaFunction{subst} relies on some auxiliary functions for
\emph{renaming} type variables and \emph{lifting} substitutions across
quantifiers. These functions are straightforward adaptations of those described
by \citet{KokkeSiekWadler20}.

>>>>

It is straightforward to prove that duality is an involution.

@@

This property is important in the rest of the formalisation so we define an
implicit rewriting rule that Agda can autonomously apply whenever possible. This
is achieved by means of the following pragma directive.\footnote{The directive
is effective provided that the option \texttt{{-}{-}rewriting} is enabled,
either globally when invoking Agda or within an \AgdaPragma{OPTIONS} pragma
directive in the module's source code.}

@@

Next we define the function \AgdaFunction{subst} that simultaneously substitutes
the type variables of a pre-type with other pre-types. In practice we will
always substitute one variable at a time, but it is technically easier to define
\AgdaFunction{subst} so that it accepts a function substituting \emph{all}
variables of a pre-type, possibly with themselves. The definition of
\AgdaFunction{subst} relies on some auxiliary functions for \emph{renaming} type
variables and \emph{lifting} substitutions across quantifiers. These functions
are straightforward adaptations of those described by \citet{KokkeSiekWadler20}.
